---
title: "Gapminder Exploration"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Author: Ray Wu

## Bringing in Rectangular data

First, we load the `gapminder` and `tidyverse` packages: 

```{r}
library(gapminder)
library(tidyverse)
```
## Smell Test: 
Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
```
As we can see, `gapminder` is stored in a list. 

What is its class?

```{r}
class(gapminder)
```

`gapminder` belongs to the classes `tibble` and `data.frame`. 

How many variables/columns?
```{r}
ncol(gapminder)
```
`gapminder` has 6 columns. 

How many rows/observations?
```{r}
nrow(gapminder)
```
There are 1704 observations in `gapminder`

Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?

```{r}
names(gapminder)
```
This can also be used to find the number of columns in a dataset

What data type is each variable?

```{r}
str(gapminder)
```
## Exploring individual & Exploring various plot types
Pick at least one categorical variable and at least one quantitative variable to explore.

categorical variable selected: continent
quantitative variable selected: lifeExp

What are possible values (or range, whichever is appropriate) of each variable?

```{r}
summary(gapminder$continent)
```

The same information, in bar chart form: 
```{r}
ggplot(gapminder, aes(continent)) + 
  geom_bar(fill = 'dark green')
```
```{r}
summary(gapminder$lifeExp)
```

When we have a continuous variable we get more information from a density plot vs. the summary statistics. 

```{r}
ggplot(gapminder, aes(lifeExp)) + 
  geom_density(bw = 0.01, fill = "orange") + 
  scale_x_log10()
```

As expected, we have a large group of countries in the 70s for life expectancy, presumably the Western (i.e. developed) countries. 

A fancier plot, illustrating the clustering of countries' life expectancy and gdp per capita by the continent. 
```{r}
plot = ggplot(gapminder, aes(gdpPercap, lifeExp)) + scale_x_log10()
plot + 
  geom_point(aes(colour = factor(continent)), size = 2)

```


What values are typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.
Feel free to use summary stats, tables, figures. We're NOT expecting high production value (yet).

Typical values of continent are the five continents: Africa, Americas, Asia, Europe, and Oceania. Note that North, Central, and South America are all mixed into one group. 

As expected, typical values of the life expectancy range from 20s to 80s. A large proportion of countries range in the 70s and higher. 


Life expectancy distribution according to continent: 

```{r}
a = ggplot(gapminder, aes(continent, lifeExp))
a + 
  geom_violin() + 
  geom_jitter(alpha=0.2)
```

## Use of filter, select, and %>%

And the following is a plot of Sudan's GDP per capital and life expectancy over time. 
```{r}
gapminder %>% 
  filter(country == 'Sudan') %>% 
  select(gdpPercap, lifeExp) %>% 
  ggplot(aes(gdpPercap, lifeExp)) + 
    geom_point() + 
    geom_path(arrow=arrow())
```


## But I want to do more!

Evaluate this code and describe the result. Presumably the analyst's intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

code: `filter(gapminder, country == c("Rwanda", "Afghanistan"))`
(Unfortunately I can't put it in an `R` chunk because then knit will have a problem with it)
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```


The problem with this code is that we do not get every row in the dataset. To get it, we use the following code: 

```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

## Additional Exercises: 

__Exercise 1__: Make a plot of `year` (x) vs `lifeExp` (y), with points coloured by continent. Then, to that same plot, fit a straight regression line to each continent, without the error bars. If you can, try piping the data frame into the `ggplot` function.

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, color=continent)) + 
  geom_point() + 
  geom_smooth(method='lm', se = FALSE)
```

__Exercise 2__: Repeat Exercise 1, but switch the _regression line_ and _geom\_point_ layers. How is this plot different from that of Exercise 1?

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, color=continent)) + 
  geom_smooth(method='lm', se = FALSE) +
  geom_point()
```

__Exercise 3__: Omit the `geom_point` layer from either of the above two plots (it doesn't matter which). Does the line still show up, even though the data aren't shown? Why or why not?

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, color=continent)) +
  geom_smooth(method='lm', se = FALSE)
```


__Exercise 4__: Make a plot of `year` (x) vs `lifeExp` (y), facetted by continent. Then, fit a smoother through the data for each continent, without the error bars. Choose a span that you feel is appropriate.
```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp)) + 
  facet_wrap(~ continent) + 
  geom_point() + 
  geom_smooth(se = F)
```

__Exercise 5__: Plot the population over time (year) using lines, so that each country has its own line. Colour by `gdpPercap`. Add alpha transparency to your liking. 

```{r}
gapminder %>% 
  ggplot(aes(year, pop)) + 
  geom_line(aes(group=country, color=gdpPercap), alpha = 0.23) + 
  scale_y_log10()
```

__Exercise 6__: Add points to the plot in Exercise 5.

```{r}
gapminder %>% 
  ggplot(aes(year, pop)) + 
  geom_line(aes(group=country, color=gdpPercap), alpha = 0.23) + 
  geom_point(alpha = 0.1) + 
  scale_y_log10() 
```